Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Validation

### 2012-12-21 paper.tex
more updates / corrections

We use the Castro code as described in \citet{castro}.  For all the runs,
the PPM reconstruction is done, using the original limiters \citep{ppm}.
The original implementation of the PPM prediction in Castro takes the
form:
\begin{equation}
q_{i+1/2,L}^{n+1/2} = q_i -
   \sum_{\nu;\lambda^{(\nu)}\ge 0} l^{(\nu)} \cdot \left [
        q_i - \mathcal{I}_+(\sigma_\nu)
       \right ] r^{(\nu)}
\end{equation}
where $q$ is the vector of primitive variables, with $q_i$
representing the average in the cell, $l$ and $r$ are the left and
right eigenvectors with eigenvalue $\lambda$.  The sum is over all the
waves that result from the characteristic structure of the problem,
but designed such that only waves moving toward the interface
contribute to the interface value, $q_{i+1/2,L}^{n+1/2}$.  Finally,
$\mathcal{I}_+(\sigma_\nu)$ is the jump carried to the right interface
of the cell by the wave $\nu$, constructed by averaging under the
parabolic interpolant in the cell all information that can reach the
interface over the timestep, with $\sigma$ representing the fraction
of the zone crossed by the wave denoted by $\nu$ during the timestep.
The reader is referred to \citet{ppmunsplit} for further defaults.

The presence of $q_i$ in this expression serves as a reference state.
The idea is that the error in the characteristic projection of the
jump (due to the nonlinearity) is minimized if we pick a suitable
reference state, since we only care about the jumps carried to the
interface over the timestep \citep{colellaglaz1985}.  The original
Castro implementation used the cell-average quantity.  Experiments
found that this is ill-behaved in the presence of strong shocks
(especially at low resolution). For the present work, we switch the
reference state to the jump carried by the fastest wave moving toward
the interface, $\mathcal{I}_+(\sigma_3)$ in this case, where the $3$
here means the wave associated with the $u + c$ eigenvalue.  This is
in agreement with \citet{ppmunsplit} (eq. 90).  This makes our
expression appear as:
\begin{equation}
q_{i+1/2,L}^{n+1/2} = \mathcal{I}_+(\sigma_3) -
   \sum_{\nu;\lambda^{(\nu)}\ge 0} l^{(\nu)} \cdot \left [
        \mathcal{I}_+(\sigma_3)  - \mathcal{I}_+(\sigma_\nu)
       \right ] r^{(\nu)}
\end{equation}

We also modify slightly the correction done to the gravitational
source terms after integrating the system in time.  In the original
Castro paper (Eqs. 19 and 20), the state after the conservative update
due to fluxes and external sources (other than gravity) was
represented with the superscript `$(2,\star)$'.  The gravitational
sources were then corrected, making them centered in time, yielding
the final state at the end of the hydrodynamics step, indicated with
the superscript `$(2)$'.  This appeared as:
\begin{eqnarray}
(\rho {\bf u})^{(2)} &=& (\rho {\bf u})^{(2,\star)} +
   \frac{\Delta t}{2} \left [ (\rho {\bf g})^{(2,\star)} - 
                              (\rho {\bf g})^{(1)} \right ]\\
(\rho E)^{(2)} &=& (\rho E)^{(2,\star)} +
   \frac{\Delta t}{2} \left [ (\rho {\bf u\cdot g})^{(2,\star)} - 
                              (\rho {\bf u\cdot g})^{(1)} \right ]
\end{eqnarray}
Here we make the subtle change of using the corrected ${\bf u}$ in the
correction to the energy.  So our new energy correction equation appears as:
\begin{equation}
(\rho E)^{(2)} = (\rho E)^{(2,\star)} +
   \frac{\Delta t}{2} \left [ (\rho {\bf u\cdot g})^{(2)} - 
                              (\rho {\bf u\cdot g})^{(1)} \right ]
\end{equation}
Both of these are second-order accurate, but we've found that the latter
is slightly better for conserving energy.

>>>>

We use the Castro code as described in \citet{castro}.  For all the runs,
the PPM reconstruction is done, using the original limiters \citep{ppm}.
The original implementation of the PPM prediction in Castro takes the
form:
\begin{equation}
q_{i+1/2,L}^{n+1/2} = q_i -
   \sum_{\nu;\lambda_i^{(\nu)}\ge 0} l_i^{(\nu)} \cdot \left [
        q_i - \mathcal{I}_+^{(\nu)}(q_i)
       \right ] r_i^{(\nu)}
\end{equation}
where $q$ is the vector of primitive variables, with $q_i$
representing the average in the cell, $l^{(\nu)}$ and $r^{(\nu)}$ are
the left and right eigenvectors with eigenvalue $\lambda^{(\nu)}$,
with $\nu$ the index of the characteristic wave of the system.  The
sum is over all the waves that result from the characteristic
structure of the problem, but designed such that only waves moving
toward the interface contribute to the interface value,
$q_{i+1/2,L}^{n+1/2}$.  Finally, $\mathcal{I}_+^{(\nu)}(q)$ is the
average under the parabolic profile of quantity $q$ of all the information
that can reach the right interface of the zone $i$ as carried by the wave
$\nu$.   The reader is referred to
\citet{ppmunsplit} for further defaults.

The presence of $q_i$ in this expression serves as a reference state.
The idea is that the error in the characteristic projection of the
jump (due to the nonlinearity) is minimized if we pick a suitable
reference state, since we only care about the jumps carried to the
interface over the timestep \citep{colellaglaz1985}.  The original
Castro implementation used the cell-average quantity.  Experiments
found that this is ill-behaved in the presence of strong shocks
(especially at low resolution). For the present work, we switch the
reference state to
\begin{equation}
\tilde{q}_L = \left \{ \begin{array}{cc}
       \mathcal{I}_+^{(+)}(q_i) & \mathrm{if~} u + c > 0 \\
       q_i                    & \mathrm{otherwise}
\end{array}
\right .
\end{equation}
where the $(+)$ superscript here means the fastest wave moving to the right
(the $u+c$ eigenvalue).   This is simply the average under the largest
portion of the parabolic profile that could possible reach the interface 
over the timestep.  This is
in agreement with \citet{ppmunsplit} (eq. 90).  This makes our
expression appear as:
\begin{equation}
q_{i+1/2,L}^{n+1/2} = \tilde{q}_L -
   \sum_{\nu;\lambda_i^{(\nu)}\ge 0} l_i^{(\nu)} \cdot \left [
        \tilde{q_i}  - \mathcal{I}^{(\nu)}_+(q_i)
       \right ] r_i^{(\nu)}
\end{equation}
This change can be used in the public version of Castro by setting
{\tt castro.ppm\_reference = 1}.

We also modify slightly the correction done to the gravitational
source terms after integrating the system in time.  In the original
Castro paper (Eqs. 19 and 20), the state after the conservative update
due to fluxes and external sources (other than gravity) was
represented with the superscript `$(2,\star)$'.  The gravitational
sources were then corrected, making them centered in time, yielding
the final state at the end of the hydrodynamics step, indicated with
the superscript `$(2)$'.  This appeared as:
\begin{eqnarray}
(\rho {\bf u})^{(2)} &=& (\rho {\bf u})^{(2,\star)} +
   \frac{\Delta t}{2} \left [ (\rho {\bf g})^{(2,\star)} - 
                              (\rho {\bf g})^{(1)} \right ]\\
(\rho E)^{(2)} &=& (\rho E)^{(2,\star)} +
   \frac{\Delta t}{2} \left [ (\rho {\bf u\cdot g})^{(2,\star)} - 
                              (\rho {\bf u\cdot g})^{(1)} \right ]
\end{eqnarray}
Here we make the subtle change of using the corrected ${\bf u}$ in the
correction to the energy.  So our new energy correction equation appears as:
\begin{equation}
(\rho E)^{(2)} = (\rho E)^{(2,\star)} +
   \frac{\Delta t}{2} \left [ (\rho {\bf u\cdot g})^{(2)} - 
                              (\rho {\bf u\cdot g})^{(1)} \right ]
\end{equation}
Both of these are second-order accurate, but we've found that the latter
is slightly better for conserving energy.  This change can be used in the
public version of Castro by setting {\tt castro.grav\_source\_type = 2}.
